Hierarchical state space models reveal dynamics of neural activity in \textit{C. elegans}

### 2018-02-26 drafts/elife/draft.tex
updates leading up to job talk

% \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
\title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
\author{Scott Linderman$^{\text{1}}$,
  Annika Nichols$^{\text{2}}$,
  David Blei$^{\text{1}}$,
  Manuel Zimmer$^{\text{2}}$,
  and
  Liam Paninski$^{\text{1}}$
  \\
  $^{\text{1}}$Columbia University,
  $^{\text{2}}$Research Institute of Molecular Pathology (IMP), Vienna Biocenter
}

@@

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a robust,
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral ``syllables'' that drive dynamics in this latent state
  space.  We find stochastic transition patterns between these
  syllables, and we see that transition probabilities are determined
  by a combination of current brain state and environmental cues.  In
  addition to quantifying neural dynamics, this probabilistic
  framework aids in neural identification---currently a laborious,
  manual task---and reveals clusters of neurons that are similarly
  tuned in latent state space.  Finally, we find a significant overlap
  between our inferred syllables and the manually-labeled states
  of~\citet{kato2015global} and~\citet{nichols2017global}, which were
  shown to correspond to different behaviors, like forward crawling,
  reversals, and turns. Our methods automatically discover and
  quantify these behaviorally-meaningful states directly from neural
  activity, yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

\textbf{Main results, in broad brushstrokes:}
\begin{itemize}
\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms. The framework is:
  \begin{itemize}
  \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
  \item \emph{robust}, ameliorating some concerns of model misspecification;
  \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
  \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
  \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
  \end{itemize}
\item We use this framework to extract low-dimensional state
  trajectories for each worm, using shared correlations across the
  population to align the trajectories and make predictions about
  unseen activity.
\item We discover clusters of similarity-tuned and highly correlated
  neurons, and we find that these neurons are functionally grouped
  according the different behaviors they are implicated in.
\item We show how our model's predictions can aid in identifying neurons (which will then drive down posterior uncertainty in the underlying dynamics of neural activity).
\item We show that brain dynamics are well-approximated as switching between discrete regimes, or ``syllables,'' each associated with simple linear dynamics.  These syllables map onto the manual segmentations of experimental experts, which were based on known behavioral implications of certain patterns of brain activity.
\item We quantify these linear dynamical regimes, show how they drive activity in clusters of neurons, and evaluate their differences from worm to worm.
\item We find that transitions between syllables are not simply Markovian; rather, the transitions depend on a combination of syllable history and latent brain state.
\item We measure how oxygen concentration (and change in oxygen concentration) differentially impact state transitions in worms of different genetic strains and developmental stages, and we find that sudden increases in oxygen elicit a characteristic, transient response regardless of genetic make-up or developmental stage, but that only certain genetic strains show a sustained response to increased oxygen levels. 
\end{itemize}

@@

@@

We analyze the data presented in~\citet{kato2015global} and
\citet{nichols2017global}, which consist of a collection of optical
recordings of calcium fluorescence in immobilized \celegans. The
recordings are 18 minutes long with a frame rate of about 3Hz.  On
average, about 100 single units are extracted from each
recording. \celegans~is genetically stereotyped and each of its 302
neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
units could be unambiguously labeled in each
recording. Fig.~\ref{fig:model}a illustrates this type of data. The
three panels represent recordings from three separate worms, and the
rows correspond to different neuron labels.  Neural activity is
measured for only a subset of neurons in each worm.

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v3/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamics model for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~Though location in continuous state space helps inform
  transitions, the current discrete state does as well. This is captured
  by the Markov transition
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

% To model this data, we build on switching linear dynamical systems
% (SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
%   ghahramani1996switching, murphy1998switching}, a type of state space
% model that views data as a projection of a low-dimensional latent
% state that switches between linear dynamical regimes.  Formally,
% let~$y_t \in \reals^N$ denote the vector of neural activity observed
% at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
% where~$x_t \in \reals^D$ is a continuous latent state
% and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
% reflecting the low-dimensional nature of neural activity. SLDS
% simplify the dynamics of the continuous latent states by introducing a
% discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
% current dynamical regime.  Each discrete regime~$k$ is associated with
% a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
% specifies which map to use in order to propagate~$x_t$ to the next
% time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
% between different linear dynamical regimes, the SLDS captures highly
% nonlinear dynamics.  We seek to infer the continuous and discrete
% latent states and simultaneously learn the linear maps, the
% distribution of the noise in the observations and the continuous
% states, and the dynamics that govern how the discrete latent states
% evolve.
To model this data, we build on switching linear dynamical systems
(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
  ghahramani1996switching, murphy1998switching}, a type of state space
model that views data as a projection of a low-dimensional latent
state that switches between linear dynamical regimes.  The expected
activity is a linear function of an underlying, continuous latent
state, illustrated as two-dimensional time series for each worm in
Fig.~\ref{fig:model}b.  We typically assume this latent state is lower
dimensional than the number of observed neurons. SLDS simplify the
dynamics of the continuous latent states by introducing discrete
latent states that indicate the current dynamical regime
(Fig.~\ref{fig:model}c).  Each discrete regime is associated with a
linear map (Fig.~\ref{fig:model}d), and the discrete state specifies
which map to use in order to propagate the continuous latent state to
the next time step.  By switching between different linear dynamical
regimes, the SLDS captures highly nonlinear dynamics.  We seek to
infer the continuous and discrete latent states and simultaneously
learn the linear maps, the distribution of the noise in the
observations and the continuous states, and the dynamics that govern
how the discrete latent states evolve.

% Recurrent model
Standard SLDS treat the discrete latent states as a simple Markov
process: the next state depends only on the current state. However,
this neglects the more nuanced ways in which discrete and continuous
states may co-evolve. For example, the next discrete state may depend
on both the current discrete state and the current location in
continuous latent space; intuitively, different dynamical regimes may
be employed in different regions of continuous latent space. These
types of dependencies are captured by what are variously called
hybrid~\citep{paoletti2007identification},
augmented~\citep{barber2006expectation}, or \emph{recurrent} state
space models~\citep{linderman2017recurrent}. The maps in
Fig.~\ref{fig:model}d illustrate how these dependencies may modulate
the probability of different dynamics depending on the current
location in continuous latent space. We employ these recurrent models
to learn dynamical regimes that are selectively employed as a
function of current continuous brain state.

% Robust model
While recurrent SLDS can approximate highly nonlinear systems, they
are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
ameliorate this concern by lessening the penalty on observations that
do not quite match the dynamics of the current regime.  We achieve
this robustness by replacing the standard Gaussian noise model with a
multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
of this distribution Fig.~\ref{fig:model}e compares a multivariate-t
distribution to a Gaussian with the same mean; though they have
the same covariance shape, the tails of the robust model fall off
more slowly than those of the Gaussian, allowing for outliers and a
degree of model misspecification.

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activity may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

% Inference
We fit the the canonical and per-worm dynamics libraries, the
parameters of the dynamics and observation noise models, and the
discrete and continuous latent states with approximate Bayesian
inference as described in the Supplementary Material. Briefly, we
separate the problem into two steps: learning the continuous latent
states and the mapping from continuous latent states to observations,
then learning the dynamics libraries that govern the continuous latent
states and the corresponding discrete state sequences.  Both steps are
performed with a block Gibbs sampler, leveraging the structure of the
probabilistic model to perform efficient conditional updates of the
various parameters.

@@

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  We
fit our model directly to the first-order differences in calcium
fluorescence; our only preprocessing step is to correct for
photo-bleaching. The green lines in Fig.~\ref{fig:smoothing} show a
three minute window of bleaching-corrected, differenced calcium traces
for two of the five worms in the~\citet{kato2015global} dataset. Of
these five worms, Worm 1 has the fewest labeled neurons (24 neurons)
and worm 5 has the most (48 neurons). Across the five worms, 61 unique
neuron labels were identified in at least one worm.

By aggregating information across worms, the model learns correlation
structure that enables it to smooth noisy traces and predict the
activity of unobserved neurons.  The smoothed and predicted activity
is shown as black lines in Fig.~\ref{fig:smoothing}. For example,
even though \textsf{SMDVR} is not labeled in worm 1, the model
learns how its activity covaries with that of other neurons
by leveraging information from other worms, like worm 5, in which
both~\textsf{SMDVR} and~\textsf{SMDVL} are labeled.  Since these
two neurons are highly correlated, the model can make strong predictions
about the activity of~\textsf{SMDVR} in worm 1 on the basis of the
activity of~\textsf{SMDVL}.

The degree to which the smoothed activity matches the observed
activity is governed by the dimensionality of the latent state space.
We estimate this dimensionality with cross validation, reserving the
last 20\% of the data (3.6 min) for testing and measuring the marginal
log likelihood of the held-out data as we increase the latent
dimensionality. We find that the test likelihood is maximized with a
10-dimensional latent state, and that variances differ substantially
between worms.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure3} 
\caption{ \textit{The mapping from continuous latent states to neural
    activity reveals clusters of functionally similar neurons.}
  \textbf{a.} The emission matrix is a linear map from latent states;
  the rows can be seen as ``tuning vectors'' that specify how each
  neuron responds to changes in the latent state.  These tuning
  vectors cluster into discrete groups (separated by black lines),
  indicating that there are subpopulations of similarly tuned, and
  hence highly correlated, neurons. Neurons are ordered as in
  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
  tuning vectors supports this clustering and shows differences
  between subpopulations (clusters have been sorted to make this as
  close to diagonal as possible).  \textbf{c.}~The tuning similarity
  reflects correlations between neurons, but since it is learned as
  part of a hierarchical model, it predicts similarities for all pairs
  of neurons, even pairs that are never simultaneously recorded (gray,
  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
  neurons from each cluster, as specified by WormAtlas.  Neurons in
  the same cluster are not necessarily close to one another in space,
  though some clusters are spatially localized. Black bars denote
  approximate locations of the anterior (AG), dorsal (DG), ventral
  (VG) and retrovesicular ganglia.}
\label{fig:clustering}
\end{figure}

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping is given by a matrix with
rows corresponding to neurons and columns corresponding to latent
dimensions, shown in Fig.~\ref{fig:clustering}a.  Each row can be seen
as a \emph{tuning vector} that specifies how the corresponding neuron
responds to each dimension of the latent state.  This matrix reveals
groups of similarly tuned neurons. To capture this, we used k-means to
cluster the neurons based on their tuning vectors and found that 12
clusters adequately captures the diversity of the population.  Using
fewer clusters leads to qualitatively similar groups but omits some of
the finer within-group differences; allowing more clusters eventually
leads to groups of lateral pairs (like \textsf{ASKL} and \textsf{ASKR}
in a.).  The neurons are ordered as in Fig.~\ref{fig:smoothing}, and
clusters are separated by black lines.

@@

What can explain this clustering?  We first look at the spatial
locations of neurons in each cluster, under the hypothesis that
clusters could be trivially explained by an inability to resolve
differences in activity between nearby neurons.
Fig.~\ref{fig:clustering}d shows the locations of all 61 labeled
neurons, using the positions from WormAtlas\todo{cite} along the
head-tail axis of the worm's body.  The size of the dot indicates the
number of neurons at that location; the bars below denote the
approximate location of the anterior (AG), dorsal (DG), ventral (VG)
and retrovesicular (RVG) ganglia.  Some cluster (e.g. cluster 6) are
spatially localized, but many clusters contain neurons from multiple
head ganglia.  This suggests that location alone cannot explain the
similarity between tuning vectors.

\todo[inline]{Break down clustes by type (sensory/inter/motor
  neuron).}

\todo[inline]{Ask Manuel/Annika to label functional role of neurons.}

@@

To accurately model these dynamics, the changes must occur at precise
locations in continuous latent space.  Namely,
Fig.~\ref{fig:syllables}a shows that the loops tend to start and end
at the origin, and we rarely see transitions when the continuous
latent state is far from zero. This manifests in signature duration
distributions for the various states.  Fig.~\ref{fig:recurrent}c plots
these distributions for each of the eight states.  Superimposed on
top, we show the predicted distributions with the recurrent SLDS
(solid black lines) compared with the predictions of the standard,
Markovian SLDS (dotted black lines). The Markovian model knows nothing
of the continuous latent state and implies a geometric duration
distribution with monotonically decreasing probability. By contrast,
the recurrent model learns to transition near the origin, which
gives rise to characteristic duration distributions that closely
match the empirical distributions.

>>>>

% \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
% \title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
\title{Hierarchical state space models reveal dynamics of neural activity in \textit{C. elegans}}
\author{Scott Linderman$^{\text{1}}$,
  Annika Nichols$^{\text{2}}$,
  David Blei$^{\text{1}}$,
  Manuel Zimmer$^{\text{2}}$,
  and
  Liam Paninski$^{\text{1}}$
  \\
  $^{\text{1}}$Columbia University,
  $^{\text{2}}$Research Institute of Molecular Pathology (IMP), Vienna Biocenter
}

@@

\begin{abstract}
  Recent advances in neural recording technologies enable
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral ``syllables'' that drive dynamics in this latent state
  space.  We find stochastic transition patterns between these
  syllables, and we see that transition probabilities are determined
  by both current brain state and environmental cues.  In
  addition to quantifying neural dynamics, this probabilistic
  framework aids in neural identification---currently a laborious,
  manual task---and reveals clusters of neurons that are similarly
  tuned in latent state space.  Finally, we find a significant overlap
  between our inferred syllables and the manually-labeled states
  of~\citet{kato2015global} and~\citet{nichols2017global}, which were
  shown to correspond to different behaviors, like forward crawling,
  reversals, and turns. Our methods automatically discover and
  quantify these states directly from neural
  activity, yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

\textbf{Main results, in broad brushstrokes:}
\begin{itemize}
\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms.
  % The framework is:
  % \begin{itemize}
  % \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
  % \item \emph{robust}, ameliorating some concerns of model misspecification;
  % \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
  % \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
  % \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
  % \end{itemize}
\item We use this framework to extract low-dimensional state
  trajectories for each worm, using shared correlations across the
  population to align the trajectories and make predictions about
  unseen activity.
\item We discover clusters of similarity-tuned and highly correlated
  neurons, and we find that these neurons are functionally grouped
  according the different behaviors they are implicated in.
\item We show how our model's predictions can aid in identifying neurons (which will then drive down posterior uncertainty in the underlying dynamics of neural activity).
\item We show that brain dynamics are well-approximated as switching between discrete regimes, or ``syllables,'' each associated with simple linear dynamics.  These syllables map onto the manual segmentations of experimental experts, which were based on known behavioral implications of certain patterns of brain activity.
\item We quantify these linear dynamical regimes, show how they drive activity in clusters of neurons, and evaluate their differences from worm to worm.
\item We find that transitions between syllables are not simply Markovian; rather, the transitions depend on a combination of syllable history and latent brain state.
\item We measure how oxygen concentration (and change in oxygen concentration) differentially impact state transitions in worms of different genetic strains and developmental stages, and we find that sudden increases in oxygen elicit a characteristic, transient response regardless of genetic make-up or developmental stage, but that only certain genetic strains show a sustained response to increased oxygen levels. 
\end{itemize}

@@

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v4/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamics model for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state, or \emph{syllable}.
  \textbf{d.}~These syllables index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~Though location in continuous state space helps inform
  transitions, the current syllable does as well. This is captured
  by a Markov transition matrix, which is modulated by location in
  continuous state space.
  \textbf{f.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{g.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{h.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

We analyze the data presented in~\citet{kato2015global} and
\citet{nichols2017global}, which consist of a collection of optical
recordings of calcium fluorescence in immobilized \celegans. The
recordings are 18 minutes long with a frame rate of about 3Hz.  On
average, about 100 single units are extracted from each
recording. \celegans~is genetically stereotyped and each of its 302
neurons has a unique label (e.g. \textsf{AVAL}). In
the~\citet{kato2015global} dataset, between 24 and 48 neurons could be
unambiguously labeled in each recording, and 61 unique labels were
found in at least one of the five worms.  Of these, 13 labels were
found in only worm, 14 were found in two, 6 were found in three, 13 in
four, and 15 were found in all five
worms. The~\citet{nichols2017global} dataset is similar, but it
contains 44 worms from two different genetic strains and two different
developmental stages.  Across these 44 worms, 73 neuron labels were
identified at least once. Fig.~\ref{fig:model}a illustrates this type
of data. The three panels represent recordings from three separate
worms, and the rows correspond to different neuron labels. Since
calcium fluorescence is highly autocorrelated from one time bin to the
next, we model the first-order differences in fluorescence instead.
Most importantly, this neural activity is measured for only a subset
of labeled neurons in each worm.

% To model this data, we build on switching linear dynamical systems
% (SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
%   ghahramani1996switching, murphy1998switching}, a type of state space
% model that views data as a projection of a low-dimensional latent
% state that switches between linear dynamical regimes.  Formally,
% let~$y_t \in \reals^N$ denote the vector of neural activity observed
% at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
% where~$x_t \in \reals^D$ is a continuous latent state
% and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
% reflecting the low-dimensional nature of neural activity. SLDS
% simplify the dynamics of the continuous latent states by introducing a
% discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
% current dynamical regime.  Each discrete regime~$k$ is associated with
% a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
% specifies which map to use in order to propagate~$x_t$ to the next
% time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
% between different linear dynamical regimes, the SLDS captures highly
% nonlinear dynamics.  We seek to infer the continuous and discrete
% latent states and simultaneously learn the linear maps, the
% distribution of the noise in the observations and the continuous
% states, and the dynamics that govern how the discrete latent states
% evolve.
We model this data with an extended version of a classical switching
linear dynamical system (SLDS) \citep{chang1978state,
  ackerson1970state, hamilton1990analysis, ghahramani1996switching,
  murphy1998switching}, a type of state space model that treats data
as a projection of a low-dimensional latent state that switches
between linear dynamical regimes.  The expected activity is a linear
function of an underlying, continuous latent state, illustrated as
two-dimensional time series for each worm in Fig.~\ref{fig:model}b.
We typically assume this latent state is lower dimensional than the
number of observed neurons. SLDS simplify the dynamics of the
continuous latent states by introducing discrete latent states, or
\emph{syllables}, that indicate the current dynamical regime
(Fig.~\ref{fig:model}c).  Each syllable is associated with a linear
map (Fig.~\ref{fig:model}d), and the discrete syllable specifies which
map to use in order to propagate the continuous latent state to the
next time step.  By switching between different linear dynamical
regimes, the SLDS captures highly nonlinear dynamics.  We seek to
infer both the continuous latent states and discrete syllables and
simultaneously learn the linear dynamics libraries, the distribution
of the noise in the observations and the continuous states, and the
transition structure that governs how the discrete syllables evolve
over time.

% Recurrent model
Standard SLDS treat the discrete syllables as a simple Markov
process: the next syllable depends only on the current syllable according
to a transition matrix, as shown in~Fig.~\ref{fig:model}e. However,
this neglects the more nuanced ways in which discrete and continuous
states may co-evolve. For example, the next discrete syllable may depend
on both the current syllable and the current location in
continuous latent space; intuitively, different dynamical regimes may
be employed in different regions of continuous latent space. These
types of dependencies are captured by what are variously called
hybrid~\citep{paoletti2007identification},
augmented~\citep{barber2006expectation}, or \emph{recurrent} state
space models~\citep{linderman2017recurrent}. The maps in
Fig.~\ref{fig:model}d illustrate how these dependencies may modulate
the probability of different dynamics depending on the current
location in continuous latent space. We employ these recurrent models
to learn dynamical regimes that are selectively employed as a
function of current continuous brain state.

% Robust model
While recurrent SLDS can approximate highly nonlinear systems, they
are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
ameliorate this concern by reducing the penalty placed on observations
that do not quite match the dynamics of the current regime.  We
achieve this robustness by replacing the standard Gaussian noise model
with a multivariate-t distribution~\citep{lange1989robust}.  The heavy
tails of this distribution Fig.~\ref{fig:model}f show a multivariate-t
distribution. Its tails decay more slowly than those of the
Gaussian, allowing for outliers and a degree of model
misspecification.

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activity may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}h.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}g) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

% Inference
We fit the the canonical and per-worm dynamics libraries, the
parameters of the dynamics and observation noise models, and the
discrete and continuous latent states with approximate Bayesian
inference as described in Section~\ref{sec:slds}. Briefly, we separate
the problem into two steps: learning the continuous latent states and
the mapping from continuous latent states to observations, then
learning the dynamics libraries that govern the continuous latent
states and the corresponding discrete syllable sequences.  This
separation is justified when the observed neural activity, combined
with a weak prior distribution to prefer smooth latent states,
provides sufficient information to constrain the continous latent
state sequence.  Having obtained these latent states, inferring the
discrete syllables is equivalent to performing inference in an
autoregressive hidden Markov model. Both stages are performed with a
block Gibbs sampler, leveraging the conditionally conjugate structure
of the probabilistic model to perform efficient updates of the states
and parameters.

Section~\ref{sec:slds} provides the technical details of our
probabilistic model and our algorithm for inferring the latent states
and syllables and learning the model parameters.
Fig.~\ref{fig:graphical_model} summarizes the probabilistic framework
in the form of a probabilistic graphical model.

@@

The parameters of the learned model offer insights into the structure
of neural activity in the head ganglia that can only be obtained by
pooling information across many partial recordings.  We fit our model
to the first-order differences in calcium fluorescence with minimal
preprocessing; we simply correct for photo-bleaching, z-score the
resulting traces, and compute the differences between consecutive
frames. The green lines in Fig.~\ref{fig:smoothing} show a three
minute window of bleaching-corrected, differenced calcium traces for
two of the five worms in the~\citet{kato2015global} dataset. Of these
five worms, worm 1 has the fewest labeled neurons (24 neurons) and
worm 5 has the most (48 neurons). Across the five worms, 61 unique
neuron labels were identified in at least one worm.

By aggregating information across worms, the model learns correlation
structure that enables it to smooth the noisy traces and predict the
activity of unlabeled neurons.  The smoothed and predicted activity
are shown as black lines in Fig.~\ref{fig:smoothing}. For example,
even though \textsf{SMDVR} is not labeled in worm 1, the model
learns how its activity covaries with that of other neurons
by leveraging information from other worms, like worm 5, in which
both~\textsf{SMDVR} and~\textsf{SMDVL} are labeled.  Since these
two neurons are highly correlated, the model makes strong predictions
about the activity of~\textsf{SMDVR} in worm 1 on the basis of the
activity of~\textsf{SMDVL} and other the other observed neurons.

The degree to which the smoothed activity matches the observed
activity is governed by the dimensionality of the latent state space.
We estimate this dimensionality with cross validation, reserving the
last 20\% of the data ($\approx$3.6 min) for testing and measuring the
marginal log likelihood of the held-out data as we increase the latent
dimensionality. We find that the test likelihood is maximized with a
10-dimensional latent state, and that performance improves
substantially after allowing neurons different observation variances
in each worm. This may account for differences in fluorescence level,
for example.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure3} 
\caption{ \textit{The mapping from continuous latent states to neural
    activity reveals clusters of functionally similar neurons.}
  \textbf{a.} The emission matrix is a linear map from latent states;
  the rows can be seen as ``tuning vectors'' that specify how each
  neuron responds to changes in the latent state.  These tuning
  vectors cluster into discrete groups (separated by black lines),
  indicating that there are subpopulations of similarly tuned, and
  hence highly correlated, neurons. Neurons are ordered as in
  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity
  between tuning vectors supports this clustering and shows
  differences between subpopulations (clusters have been sorted to
  make this as close to diagonal as possible).  \textbf{c.}~The
  empirical correlation matrix between neurons mimics the similarity
  between tuning vectors, but has a substantial number of missing
  entries (gray, off-diagonal boxes) for pairs that were never
  simultaneously recorded .  \textbf{d.}~We show the spatial locations
  of neurons from each cluster, as specified by WormAtlas.  Neurons in
  the same cluster are not necessarily close to one another in space,
  though some clusters are spatially localized. Black bars denote
  approximate locations of the anterior (AG), dorsal (DG), ventral
  (VG) and retrovesicular ganglia.  \textbf{e.}~Clusters 2 and 7
  contain only sensory (S) neurons, but the other clusters contain
  neurons with a mix of interneurons (I), motor (M) and polymodal (P) neurons, and
  neurons of unknown (U) type. \textbf{f.}~However, these clusters of
  interneurons and motor neurons are functionally grouped, according
  to expert labels. }
\label{fig:clustering}
\end{figure}

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping is given by a matrix with
rows corresponding to neurons and columns corresponding to latent
dimensions, shown in Fig.~\ref{fig:clustering}a.  Each row can be seen
as a \emph{tuning vector} that specifies how the corresponding neuron
responds to each dimension of the latent state.  This matrix reveals
groups of similarly tuned neurons. To capture these groups, we used
k-means to cluster the neurons based on their tuning vectors and found
that 12 clusters adequately captures the diversity of the population.
Using fewer clusters leads to qualitatively similar groups but omits
some of the finer within-group differences; allowing more clusters
eventually leads to groups of lateral pairs (like \textsf{ASKL} and
\textsf{ASKR} in Fig.~\ref{fig:clustering}a).  The neurons are ordered
as in Fig.~\ref{fig:smoothing}, and clusters are separated by black
lines.

@@

What can explain this clustering?  We first look at the spatial
locations of neurons in each cluster, under the hypothesis that
clusters could be trivially explained by an inability to resolve
differences in activity between nearby neurons.
Fig.~\ref{fig:clustering}d shows the locations of all 61 labeled
neurons, using the positions from WormAtlas \citep{wormatlas} along the
head-tail axis of the worm's body.  The size of the dot indicates the
number of neurons at that location; the bars below denote the
approximate location of the anterior (AG), dorsal (DG), ventral (VG)
and retrovesicular (RVG) ganglia.  Some cluster (e.g. cluster 6) are
spatially localized, but many clusters contain neurons from multiple
head ganglia.  This suggests that location alone cannot explain the
similarity between tuning vectors.

Next, we considered the types of cells contained in each cluster.
Fig.~\ref{fig:clustering}e shows the composition of each cluster in
terms of sensory (S), inter (I), motor (M), polymodal (P) and unknown
(U) neuron types.  Clusters 2 and 7 consist of only sensory neurons,
cluster 3 contains mostly interneurons, and clusters 8 and 11 are
primarily motor neurons, but the other clusters are composed of a mix
of inter and motor neurons. Thus, cell type alone appears insufficient to
explain these clusters as well.

@@

Though these syllables were inferred from neural activity alone, the
transition structure---forward and reverse crawling linked by dorsal
and ventral turns---mimics the natural behavior of freely crawling
worms.  This point was made by~\citet{kato2015global} based on their
manual segmentation of neural activity; here we see that this finding
holds given our automatic segmentation as well.  Moreover, we obtain
further insights by studying the precise transition structure that
links these syllables together according to the current syllable
as well as the current continuous latent brain state.
